Cluster Dynamics with HETDEX at z < 0.5 - I: Simulated Performance, Mass Distribution and Limits

### 2016-03-24 ms.tex
update with collaborators comments

The Hobby Eberly Telescope Dark Energy eXperiment (HETDEX; \citealt{Hill2008}) is a trailblazing effort to observe high-redshift large scale structures using cutting edge wide-field integral field unit (IFU) spectrographs. Designed to probe the evolution of the dark energy equation of state etched onto high redshift ($z>2$) galaxies by the Baryon Acoustic Oscillations \citep{Eisenstein2005} in the first moments of the universe, the survey will observe two fields for a total of 420 \degsq\ (300 \degsq, Spring field and 120 \degsq, Fall field). Tuned to find \lya\ emitting (LAE) galaxies at $1.9<z<3.5$, HETDEX expects to find 800,000 LAEs, and more than one million [\ion{O}{ii}] emitting galaxies at $z<0.5$ masquerading as high-redshift galaxies \citep{Acquaviva2014}.

@@

Connecting the dynamical properties derived from spectroscopy to the properties inferred from other studies insures the greatest impact on future work. HETDEX overlaps with the Sloan Digital Sky Survey (SDSS; \citealt{Blanton2001a}), SDSS stripe 82 \citep{Annis2014}, DES, the \emph{Spitzer}/HETDEX Exploratory Large-Area Survey (SHELA; Papovich \etal\ 2016, ApJS, submitted), and the \emph{Spitzer} IRAC Equatorial Survey (SpIES; Timlin \etal\ 2016, ApJ, submitted). While the potential dataset is very rich, two large issues remain.

@@

We begin with the perfect knowledge observations. These observations are of the same clusters as the targeted observations but without any observational limits. The cluster masses predicted by Equation \ref{eq:power law} gives the following results. Across all cluster masses, we find $\mu_{bias} = 0.15\pm{0.24}$ dex. The scatter in recovered masses can be attributed to both physical and numerical effects. The presence of any in-falling matter onto lower mass clusters can introduce a significant amount of substructure, which increases the observed LOSVD increasing the predicted mass \citeeg{Ntampaka2015}. Also, as the number of cluster galaxies decreases the LOSVD PDF is poorly sampled leading to poorly recovered cluster masses due to numerical effects. The masses presented here are recovered using the best possible conditions, where we have perfect knowledge of the cluster membership. In reality, the mass recovery levels presented in this section represent an upper bound (the best) on the accuracy achievable through this method.

@@

 A few caveats apply to the numbers given in Table \ref{tbl:mass bias}. While we provide corrections for cluster masses above $10^{15}$, they are estimated from only a handful of objects, and do not constitute a representative sample of clusters. On the oposite end of the cluster mass spectrum, survey observations of low mass clusters are incomplete for two potential reasons. There are very few, if any, clusters detected with survey observations below $10^{12}$. Missing corrections above $10^{12}$ in the probability based methods are due to low number statstics and an inability to sample the cluster mass PDF well enough to assign cluster mass predictions.

@@

\subsection{Impact on Cluster Cosmology}
\editorial{This is a place holder section. One of the things missing to bring this work full circle is a discussion about how the uncertainties in cluster masses impact the derived cosmological parameters. I am currently doing a literature review looking for references that discuss how the mass errors impact the derived parameters. Here we want to be able to say ``so and so show that errors on clusters masses need to be better than blah to recover $\sigma_8$ and $\Omega_m$ to better than blah, which is useful for blah. We show that with XXXX kind of data spectroscopic surveys like HETDEX can achieve this but only with YYYY kind of data quality.'' If it isn't going to do that well, then how much constraint could HETDEX/HETDEX-like surveys put on cosmological parameters? How good would they have to get to test gravity? (or the dark energy equation of state)}

>>>>

The Hobby Eberly Telescope Dark Energy eXperiment (HETDEX; \citealt{Hill2008}) is a trailblazing effort to observe high-redshift large scale structures using cutting edge wide-field integral field unit (IFU) spectrographs. Designed to probe the evolution of the dark energy equation of state etched onto high redshift ($z>2$) galaxies by the Baryon Acoustic Oscillations (BAO) \citep{Eisenstein2005} in the first moments of the universe, the survey will observe two fields for a total of 420 \degsq\ (300 \degsq, Spring field and 120 \degsq, Fall field). Tuned to find \lya\ emitting (LAE) galaxies at $1.9<z<3.5$, HETDEX expects to find 800,000 LAEs, and more than one million [\ion{O}{ii}] emitting galaxies at $z<0.5$ masquerading as high-redshift galaxies \citep{Acquaviva2014}.

@@

Connecting the dynamical properties derived from spectroscopy to the properties inferred from other studies insures the greatest impact on future work. HETDEX overlaps with the Sloan Digital Sky Survey (SDSS; \citealt{Blanton2001a}), SDSS stripe 82 \citep{Annis2014}, DES, the \emph{Spitzer}/HETDEX Exploratory Large-Area Survey (SHELA; \citealt{Papovich2016}), and the \emph{Spitzer} IRAC Equatorial Survey (SpIES; Timlin \etal\ 2016, ApJ, submitted). While the potential dataset is very rich, two large issues remain.

@@

We begin with the perfect knowledge observations. These observations are of the same clusters as the targeted observations but without any observational limits. The cluster masses predicted by Equation \ref{eq:power law} gives the following results. Across all cluster masses, we find $\mu_{bias} = 0.15\pm{0.002}$ dex. The scatter in recovered masses can be attributed to both physical and numerical effects. The presence of any in-falling matter onto lower mass clusters can introduce a significant amount of substructure, which increases the observed LOSVD increasing the predicted mass \citeeg{Ntampaka2015}. Also, as the number of cluster galaxies decreases the LOSVD PDF is poorly sampled leading to poorly recovered cluster masses due to numerical effects. The masses presented here are recovered using the best possible conditions, where we have perfect knowledge of the cluster membership. In reality, the mass recovery levels presented in this section represent an upper bound (the best) on the accuracy achievable through this method.

@@

 A few caveats apply to the numbers given in Table \ref{tbl:mass bias}. While we provide corrections for cluster masses above $10^{15}$ \msol, they are estimated from only a handful of objects, and do not constitute a representative sample of clusters. On the oposite end of the cluster mass spectrum, survey observations of low mass clusters are incomplete for two potential reasons. There are very few, if any, clusters detected with survey observations below $10^{12}$ \msol. Missing corrections above $10^{12}$ \msol\ in the probability based methods are due to low number statstics and an inability to sample the cluster mass PDF well enough to assign cluster mass predictions.

\subsubsection{Significance of Cosmology Choice}
The underlying simulation used to produce the Buzzard catalogs, as with every simulation, makes specific assumptions about the cosmology used and how the galaxies are populated into the dark matter halos. When using simulation data to train ML methods, we incorporate all of those assumptions into the learned feature associations. This could lead to a notable bias when applying the ML method to data (real or simulated) which has significantly different cosmological assumptions.

To investigate this, we select cluster halo and galaxy catalogs from the Millennium simulation \citep{Springel2005b} obtained via querying the Millennium online database\footnote{http://gavo.mpa-garching.mpg.de/Millennium/} \citep{Lemson2006}. The Millennium simulation tracks $2160^3$ dark matter particles of $8.6\times 10^8 ~h^{-1}$ \Msol\ inside a comoving 500 $(h^{-1} \mathrm{Mpc})^3$ box from $z=127$ to 0. It adopts a flat cosmological model based of the values derived from the Two-degree Field Galaxy Redshift Survey \citep{Colless2001} and the first year data of the \emph{Wilkinson Microwave Anisotropy Probe} (WMAP; \citealt{Spergel2003}): $\Omega_\Lambda = 0.75$, $\Omega_M = 0.25$, $\sigma_8 = 0.9$, $n_s = 1$ and $H_0= 73$ \kms \mpc.

The halo catalogs are created using a friends-of-friends algorithm \citeeg{Turner1976} with linking length, $b=0.2$. The halos are populated with galaxies using a semi-analytical model described in \cite{Croton2006, DeLucia2006, DeLucia2007a}.

We select 4806 clusters, comprised of 623663 galaxies, at $0.02 < z < 0.5$ and $M > 10^{13}$ \Msol, and apply the same data processing as with the Buzzard galaxies. We begin by assigning each galaxy an [\ion{O}{ii}] flux value (see Section \ref{sec: oii luminosity}), and ``observe'' each galaxy using realistic (see Section \ref{sec: observations}) observational limits. After recovering 3750 clusters which have at least five galaxies observed, we calculate the LOSVD of each cluster as in Section \ref{sec: LOSVD}.

Cluster masses are assigned using the ML methods (see Section \ref{sec:machine learning method}) using two different training sets. We use the full set of clusters detected in the Buzzard catalogs (14000 clusters with $M > 10^{11}$ \msol) and a training-testing split of the Millennium clusters.

\begin{figure} 
	\includegraphics[width=\columnwidth]{figures/millBuzzComparison.pdf} 
	\caption{Mass predictions for the power law scaling relation (Equation~\ref{eq:power law}) and the ML based technique with different input features as a function of true cluster mass. The bottom row of panels shows the fractional error (Equation~\ref{eq: fractional error}) also as a function of true cluster mass. The solid black line shows the 1:1 relation. The solid, colored line is the median predicted mass for the targeted observing, and the colored, dashed line is the median recovered mass for the HETDEX-like observations. The shaded regions represent the 68\% scatter around the median values.} \label{fig: mill buzz comparison} 
\end{figure}

The predicted cluster masses for the 4000 millennium cluster are very similar regardless of the training set used. I've attached a plot that shows the probability of the predicted masses using the two different training sets. I also computed the mean squared error (MSE) of each set, compared to the *true* cluster mass. I like the MSE because it takes into account both the scatter and the bias. For the millennium clusters I found MSE $= 0.015$, and for the Buzzard clusters MSE $= 0.029$. Smaller numbers are better, and I think the difference is because I have chosen not to exclude clusters with $Msol < 1e13$ from the Buzzard catalogs when training. That allows the ML method to assign clusters below that which ups the MSE.

I'm hoping that Millennium is different enough from Buzzard that this will address your concern. Another option would be just to include training samples from a few different mocks. Or, we could derive masses using different training sets and then marginalize over them to get a final predicted mass.

@@

\subsection{Impact on Cluster Cosmology}
% \editorial{This is a place holder section. One of the things missing to bring this work full circle is a discussion about how the uncertainties in cluster masses impact the derived cosmological parameters. I am currently doing a literature review looking for references that discuss how the mass errors impact the derived parameters. Here we want to be able to say ``so and so show that errors on clusters masses need to be better than blah to recover $\sigma_8$ and $\Omega_m$ to better than blah, which is useful for blah. We show that with XXXX kind of data spectroscopic surveys like HETDEX can achieve this but only with YYYY kind of data quality.'' If it isn't going to do that well, then how much constraint could HETDEX/HETDEX-like surveys put on cosmological parameters? How good would they have to get to test gravity? (or the dark energy equation of state)}

Galaxy clusters are sensitive probes of $\Omega_m$, the total mass ($\Omega_b + \Omega_c$) density in the universe, through tracing the peaks in the universal matter density often referred to as the power spectrum of matter density fluctuations or the matter power spectrum and $\sigma_8h^{-1}$, the normalization of the power spectrum, by the comparison of the number density of observed halos to that predicted in cosmological models. Although, in reality, one measures $\sigma_8h^{-1}\Omega_m^q$, where the value of $q$ depends on the masses and redshifts of the halos considered.

To get a sense of how well HETDEX will be able to constrain cosmological parameters we follow the discussion of \cite{Weinberg2013}, and begin with a few simplifying assumptions. While sensitive to $\Omega_m$, the number density of clusters do not necessarily provide the strongest contraint, so we assume that some combination of other data sets (\eg, CMB, BAO, supernovae, WL, etc.) will constrain $\Omega_m$ to high precision.

To estimate the error associated with a measurement of $\sigma_8h^{-1}$ (which \citealt{Weinberg2013} refer to as $\sigma_{11,abs}$), \cite{Weinberg2013} consider two sources of uncertainty, the systematic uncertainties in cluster mass calibration and the statistical uncertainty in the observed number density of clusters. The authors combine these two uncertainties though (their Eq. 141):
\begin{equation}
	\Delta~\mathrm{ln}~\sigma_{11,abs}(z) = q \times \mathrm{max}[\Delta~\mathrm{ln}~M,\alpha^{-1}\Delta\mathrm{ln}~N]
\end{equation}
where $q$ is the degeneracy exponent between $\sigma_8h^{-1}$ and $\Omega_m$, $\Delta\mathrm{ln}M$ is the mass scale uncertainty, $\Delta\mathrm{ln}N$ is the cluster statistical uncertainty, and $\alpha$ is slope of the cumulative HMF. Using the \cite{Tinker2008} HMF \cite{Weinberg2013} estimate $q\sim0.4$, and find that for any cluster survey with more than 10-20 clusters is dominated by the uncertainty in the overall mass scale.

For a survey such as HETDEX, with a power law derived overall cluster mass biases of $>0.15$ dex (about 41\%) the uncertainty on $\sigma_8h^{-1}$ is $0.4 \times 0.41 \approx 16\%$. If we consider only clusters with masses above $10^{14}$ \Msol, with perfect knowledge, and the ML based cluster mass estimates the mass scale uncertainty is reduced to $\sim -0.02$ dex or about 5\%. This gives a uncertainty on $\sigma_8h^{-1}$ of 2\%. For clusters above $10^{14}$ \Msol, survey observations can constrain the masses to about 20\% which, in turn, constrains $\sigma_8h^{-1}$ to 8\%.

Because of the simplifying assumptions, and the superior quality of the data (no contamination, signal-to-noise issues, etc.), realistic expectations for HETDEX is to directly constrain $\sigma_8h^{-1}$ to better precision than other methods (\eg, CMB, WL, X-ray) seems bleak. For the $\sim1500$ clusters detected with survey observations to constrain $\sigma_8h^{-1}$ and to be dominated by cluster statistics alone ($\Delta\mathrm{ln}N \sim N^{-1/2}$), the absolute cluster calibration would need to be better than 2.5\%. For a fully targeted survey, about 14000 clusters, this cluster mass calibration uncertainty reduces to $>1\%$.    
\editorial{Here I want to say a word or two about how all hope isn't lost. We can still use HETDEX to reduce the scatter on the richness-mass relationship which can help other estimators get down to the levels we need to really nail down these numbers.}
